A Tidy Framework and Infrastructure to Systematically Assemble Spatio-temporal Indexes from Multivariate Data

### 2023-08-22 tidyindex.qmd
fix table margin

```{r tbl-gggi-weights}
#| tbl-cap: "Weights for the two applications of dimension reduction to compute the Global Gender Gap Index. V-weight is used to compute four new variables from the original 14. These are then equally combined to get the final index value."
#| tbl-cap-location: bottom
library(kableExtra)
gggi_weights %>%  
   mutate(variable = str_replace_all(variable, "_", " ") %>%
            str_to_sentence()) %>% 
  select(c(1, 5, 2, 6)) %>% 
  mutate(dimension = c("Economy", rep("", 4), 
                       "Education", rep("" , 3), 
                       "Health", "", 
                       "Politics", rep("", 2)),
         dim_weight = c(0.25, rep("", 4),
                        0.25, rep("", 3),
                        0.25, "",
                        0.25, rep("", 2))) %>% 
  knitr::kable(
    digits = 3, booktabs = TRUE, 
    col.names = c("Variable", "V-weight", "Dimension", 
        "D-weight"),
        linesep = c("", "", "", "", "\\hline",
                    "", "", "", "\\hline",
                    "", "\\hline",
                    "", "", "\\hline"),
      align = c("l", "r", "l", "r")
    ) %>%
  kable_styling(full_width = FALSE, ) %>%
  row_spec(0, bold = TRUE)
```

>>>>

```{r tbl-gggi-weights}
#| tbl-cap: "Weights for the two applications of dimension reduction to compute the Global Gender Gap Index. V-weight is used to compute four new variables from the original 14. These are then equally combined to get the final index value."
#| tbl-cap-location: bottom
# table cap location is not working: https://github.com/quarto-dev/quarto-cli/issues/6417
library(kableExtra)
gggi_weights %>%  
   mutate(variable = str_replace_all(variable, "_", " ") %>%
            str_to_sentence()) %>% 
  select(c(1, 5, 2, 6, 7)) %>% 
  mutate(dimension = c("Economy", rep("", 4), 
                       "Education", rep("" , 3), 
                       "Health", "", 
                       "Politics", rep("", 2)),
         dim_weight = c(0.25, rep("", 4),
                        0.25, rep("", 3),
                        0.25, "",
                        0.25, rep("", 2))) %>% 
  knitr::kable(
    digits = 3, booktabs = TRUE, 
    col.names = c("Variable", "V-weight", "Dimension", "D-weight", "weight"),
    linesep = c("", "", "", "", "\\hline",
                "", "", "", "\\hline",
                "", "\\hline",
                "", "", "\\hline"),
      align = c("l", "r", "l", "r", "r")
    ) %>%
  kable_styling(font_size = 9) %>%
  row_spec(0, bold = TRUE) %>% 
  column_spec(1, width = "18em") %>% 
  column_spec(3, width = "4em") %>% 
  column_spec(c(2, 4, 5), width = "5em")
```

### 2023-08-22 tidyindex.tex
fix table margin

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
Indexes are useful for summarizing multivariate information into single
metrics for monitoring, communicating, and decision-making. While most
work has focused on defining new indexes for specific purposes, more
attention needs to be directed towards making it possible to understand
index behavior in different data conditions, and to determine how their
structure affects their values and variation in values. Here we discuss
a modular data pipeline recommendation to assemble indexes. It is
universally applicable to index computation and allows investigation of
index behavior as part of the development procedure. One can compute
indexes with different parameter choices, adjust steps in the index
definition by adding, removing, and swapping them to experiment with
various index designs, calculate uncertainty measures, and assess
indexes' robustness. The paper presents three examples to illustrate the
pipeline framework usage: comparison of two different indexes designed
to monitor the spatio-temporal distribution of drought in Queensland,
Australia; the effect of dimension reduction choices on the Global
Gender Gap Index (GGGI) on countries' ranking; and how to calculate
bootstrap confidence intervals for the Standardized Precipitation Index
(SPI). The methods are supported by a new R package, called
\texttt{tidyindex}.
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep software design\sep uncertainty\sep 
decision-making
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, interior hidden, frame hidden, enhanced, sharp corners]}{\end{tcolorbox}}\fi

@@

\hypertarget{tbl-notation}{}
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.0972}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2083}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1250}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2083}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.3611}}@{}}
\caption{\label{tbl-notation}An notation overview of the input,
operation, and output of each pipeline module.}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Section
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Module
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Input
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Operation
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Output
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Section
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Module
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Input
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Operation
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Output
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
3.1 & Temporal processing & \(x_{sj}(t)\) & \(f[x_{sj}(t)]\) &
\(x^{\text{Temp}}_{sj}(t^\prime) \quad t^\prime \in D_{t^\prime}\) \\
3.2 & Spatial processing & \(x_{tj}(s)\) & \(g[x_{tj}(s)]\) &
\(x^{\text{Spat}}_{tj}(s^\prime) \quad s^\prime \in D_{s^\prime}\) \\
3.3 & Variable transformation & \(x_{j}(s; t)\) & \(T[x_j(s;t)]\) &
\(x^{\text{Trans}}_j(s;t)\) \\
3.4 & Scaling & \(x_j(s; t)\) & \([x_j(s;t) - \alpha]/\gamma\) &
\(x^{\text{Scale}}_j(s;t)\) \\
3.5 & Dimension reduction & \(\mathbf{x}(s;t)\) & \(h[\mathbf{x}(s;t)]\)
& \(\mathbf{y}(s;t) \quad \mathbf{y} \subseteq \mathbb{R}^d, d < p\) \\
3.6 & Distribution fit & \(x_j(s; t)\) & \(F[x_j(s;t)]\) &
\(P_j(s;t) \quad P(.) \in [0, 1]\) \\
3.7 & Normalising & \(x_j(s; t)\) & \(\Phi^{-1}[x_j(s; t)]\) &
\(z_j(s; t)\) \\
3.8 & Benchmarking & \(x_j(s; t)\) & \(u[x_j(s;t)]\) & \(b_j(s;t)\) \\
3.9 & Simplification & \(x_j(s; t)\) & \(v[x_j(s;t)]\) &
\(A_j(s;t) \in \{a_1, a_2, \cdots, a_j\}\) \\
\end{longtable}

@@

\centering
\begin{tabular}{lrlrr}
\toprule
\textbf{Variable} & \textbf{V-weight} & \textbf{Dimension} & \textbf{D-weight} & \textbf{weight}\\
\midrule
Labour force participation & 0.199 & Economy & 0.25 & 0.050\\
Wage equality for similar work & 0.310 &  &  & 0.078\\
Estimated earned income & 0.221 &  &  & 0.055\\
Legislators senior officials and managers & 0.149 &  &  & 0.037\\
Professional and technical workers & 0.121 &  &  & 0.030\\
\hline
Literacy rate & 0.191 & Education & 0.25 & 0.048\\
Enrolment in primary education & 0.459 &  &  & 0.115\\
Enrolment in secondary education & 0.230 &  &  & 0.058\\
Enrolment in tertiary education & 0.121 &  &  & 0.030\\
\hline
Sex ratio at birth & 0.693 & Health & 0.25 & 0.173\\
Healthy life expectancy & 0.307 &  &  & 0.077\\
\hline
Women in parliament & 0.310 & Politics & 0.25 & 0.078\\
Women in ministerial positions & 0.247 &  &  & 0.062\\
Years with female head of state & 0.443 &  &  & 0.111\\
\bottomrule
\end{tabular}
\end{table}

>>>>

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
Indexes are useful for summarizing multivariate information into single
metrics for monitoring, communicating, and decision-making. While most
work has focused on defining new indexes for specific purposes, more
attention needs to be directed towards making it possible to understand
index behavior in different data conditions, and to determine how their
structure affects their values and variation in values. Here we discuss
a modular data pipeline recommendation to assemble indexes. It is
universally applicable to index computation and allows investigation of
index behavior as part of the development procedure. One can compute
indexes with different parameter choices, adjust steps in the index
definition by adding, removing, and swapping them to experiment with
various index designs, calculate uncertainty measures, and assess
indexes' robustness. The paper presents three examples to illustrate the
pipeline framework usage: comparison of two different indexes designed
to monitor the spatio-temporal distribution of drought in Queensland,
Australia; the effect of dimension reduction choices on the Global
Gender Gap Index (GGGI) on countries' ranking; and how to calculate
bootstrap confidence intervals for the Standardized Precipitation Index
(SPI). The methods are supported by a new R package, called
\texttt{tidyindex}.
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep software design\sep uncertainty\sep 
decision-making
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, sharp corners, breakable, enhanced, borderline west={3pt}{0pt}{shadecolor}, interior hidden, boxrule=0pt]}{\end{tcolorbox}}\fi

@@

\hypertarget{tbl-notation}{}
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.0972}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2083}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1250}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2083}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.3611}}@{}}
\caption{\label{tbl-notation}An notation overview of the input,
operation, and output of each pipeline module.}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Section
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Module
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Input
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Operation
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Output
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Section
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Module
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Input
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Operation
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Output
\end{minipage} \\
\midrule()
\endhead
3.1 & Temporal processing & \(x_{sj}(t)\) & \(f[x_{sj}(t)]\) &
\(x^{\text{Temp}}_{sj}(t^\prime) \quad t^\prime \in D_{t^\prime}\) \\
3.2 & Spatial processing & \(x_{tj}(s)\) & \(g[x_{tj}(s)]\) &
\(x^{\text{Spat}}_{tj}(s^\prime) \quad s^\prime \in D_{s^\prime}\) \\
3.3 & Variable transformation & \(x_{j}(s; t)\) & \(T[x_j(s;t)]\) &
\(x^{\text{Trans}}_j(s;t)\) \\
3.4 & Scaling & \(x_j(s; t)\) & \([x_j(s;t) - \alpha]/\gamma\) &
\(x^{\text{Scale}}_j(s;t)\) \\
3.5 & Dimension reduction & \(\mathbf{x}(s;t)\) & \(h[\mathbf{x}(s;t)]\)
& \(\mathbf{y}(s;t) \quad \mathbf{y} \subseteq \mathbb{R}^d, d < p\) \\
3.6 & Distribution fit & \(x_j(s; t)\) & \(F[x_j(s;t)]\) &
\(P_j(s;t) \quad P(.) \in [0, 1]\) \\
3.7 & Normalising & \(x_j(s; t)\) & \(\Phi^{-1}[x_j(s; t)]\) &
\(z_j(s; t)\) \\
3.8 & Benchmarking & \(x_j(s; t)\) & \(u[x_j(s;t)]\) & \(b_j(s;t)\) \\
3.9 & Simplification & \(x_j(s; t)\) & \(v[x_j(s;t)]\) &
\(A_j(s;t) \in \{a_1, a_2, \cdots, a_j\}\) \\
\bottomrule()
\end{longtable}

@@

\centering\begingroup\fontsize{9}{11}\selectfont

\begin{tabular}{>{\raggedright\arraybackslash}p{18em}>{\raggedleft\arraybackslash}p{5em}>{\raggedright\arraybackslash}p{4em}>{\raggedleft\arraybackslash}p{5em}>{\raggedleft\arraybackslash}p{5em}}
\toprule
\textbf{Variable} & \textbf{V-weight} & \textbf{Dimension} & \textbf{D-weight} & \textbf{weight}\\
\midrule
Labour force participation & 0.199 & Economy & 0.25 & 0.050\\
Wage equality for similar work & 0.310 &  &  & 0.078\\
Estimated earned income & 0.221 &  &  & 0.055\\
Legislators senior officials and managers & 0.149 &  &  & 0.037\\
Professional and technical workers & 0.121 &  &  & 0.030\\
\hline
Literacy rate & 0.191 & Education & 0.25 & 0.048\\
Enrolment in primary education & 0.459 &  &  & 0.115\\
Enrolment in secondary education & 0.230 &  &  & 0.058\\
Enrolment in tertiary education & 0.121 &  &  & 0.030\\
\hline
Sex ratio at birth & 0.693 & Health & 0.25 & 0.173\\
Healthy life expectancy & 0.307 &  &  & 0.077\\
\hline
Women in parliament & 0.310 & Politics & 0.25 & 0.078\\
Women in ministerial positions & 0.247 &  &  & 0.062\\
Years with female head of state & 0.443 &  &  & 0.111\\
\bottomrule
\end{tabular}
\endgroup{}
\end{table}
